---
title: "ukbb_vqtl_nmr_tg_overview_20210908"
author: "Tim Majarian"
date: "9/8/2021"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

\newpage
\tableofcontents

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggsci)
library(pgenlibr)

```

# Resources

## QC flag information
https://biobank.ndph.ox.ac.uk/showcase/ukb/docs/nmrm_app4.pdf  
Values with QC flag equal to 1, 3, or 9 were removed.  
Sample preparation and assay information is still pending with UKBB. We should probably explore adjusting for these when the fields get approved for our application.

## QC flag coding
Coding	Meaning  
**1	Below limit of quantification**  
2	Citrate plasma  
**3	Degraded sample**  
4	High ethanol  
5	Isopropyl alcohol  
6	Low glutamine or high glutamate  
7	Medium ethanol  
8	Polysaccharides  
**9	Unknown contamination**  
10	Ethanol  

## TG Fields
23407	Total Triglycerides
23408	Triglycerides in VLDL
23409	Triglycerides in LDL
23410	Triglycerides in HDL
23487	Triglycerides in Chylomicrons and Extremely Large VLDL
23494	Triglycerides in Very Large VLDL
23501	Triglycerides in Large VLDL
23508	Triglycerides in Medium VLDL
23515	Triglycerides in Small VLDL
23522	Triglycerides in Very Small VLDL
23529	Triglycerides in IDL
23536	Triglycerides in Large LDL
23543	Triglycerides in Medium LDL
23550	Triglycerides in Small LDL
23557	Triglycerides in Very Large HDL
23564	Triglycerides in Large HDL
23571	Triglycerides in Medium HDL
23578	Triglycerides in Small HDL
23787	Triglycerides in Chylomicrons and Extremely Large VLDL, QC Flag
23794	Triglycerides in Very Large VLDL, QC Flag
23801	Triglycerides in Large VLDL, QC Flag
23808	Triglycerides in Medium VLDL, QC Flag
23843	Triglycerides in Medium LDL, QC Flag
23850	Triglycerides in Small LDL, QC Flag
23857	Triglycerides in Very Large HDL, QC Flag
23864	Triglycerides in Large HDL, QC Flag
23871	Triglycerides in Medium HDL, QC Flag
23878	Triglycerides in Small HDL, QC Flag

## Process Fields
FieldID	Field
23649	Shipment Plate
23650	Spectrometer
23651	Measurement Quality Flagged
23652	High Lactate
23653	High Pyruvate
23654	Low Glucose
23655	Low Protein

```{r, include = F}
# Fields to include
## Still don't have the process fields
tg_fields <- c(23407, 23487,23787,23410,23529,23564,23864,23536,23501,23801,23409,23571,23871,23543,23843,23508,23808,23578,23878,23550,23850,23515,23557,23857,23494,23794,23522,23408,23435,23735,23583,23883,23613,23638,23938,23618,23593,23893,23643,23943,23623,23923,23598,23898,23648,23948,23628,23928,23603,23633, 23933, 23588, 23888, 23608)
tg_baseline_fields <- paste0("f.", tg_fields, ".0.0")
process_fields <- paste0("f.", c("23650", "23649", "23651", "23652", "23653", "23654", "23655"), ".0.0")

# qc field mapping
qc_map <- c(
  'f.23487.0.0' = 'f.23787.0.0',
  'f.23494.0.0' = 'f.23794.0.0',
  'f.23501.0.0' = 'f.23801.0.0',
  'f.23508.0.0' = 'f.23808.0.0',
  'f.23543.0.0' = 'f.23843.0.0',
  'f.23550.0.0' = 'f.23850.0.0',
  'f.23557.0.0' = 'f.23857.0.0',
  'f.23564.0.0' = 'f.23864.0.0',
  'f.23571.0.0' = 'f.23871.0.0',
  'f.23578.0.0' = 'f.23878.0.0'
)

plot_fields <- c("f.23407.0.0", "f.23408.0.0", "f.23409.0.0", "f.23410.0.0", 
"f.23487.0.0", "f.23494.0.0", "f.23501.0.0", "f.23508.0.0", "f.23515.0.0", 
"f.23522.0.0", "f.23529.0.0", "f.23536.0.0", "f.23543.0.0", "f.23550.0.0", 
"f.23557.0.0", "f.23564.0.0", "f.23571.0.0", "f.23578.0.0")
plot_adj_fields <- paste0(plot_fields, "_adj")

# input files
dd.file <- "/Users/tmajaria/Documents/projects/biobanks/ukbb/data/Data_Dictionary_Showcase_downloaded_20210908.csv"
data.dir <- "/Users/tmajaria/Documents/projects/biobanks/ukbb/data/vqtl/"
res.dir <- "/Users/tmajaria/Documents/projects/biobanks/ukbb/results/vqtl/"
phe.file <- paste0(data.dir, "vqtl_phenos_EUR.csv")
exp.file <- paste0(data.dir, "ewis_phenos_EUR.csv")
var_info.file <- paste0(data.dir, "variable_info")
pc.file <- paste0(data.dir, "ewis_phenos_EUR_PCA")
whr.file <- paste0(data.dir, "EUR_sexstrat_WHRadjBMI_INT.txt")
nmr.file <- paste0(data.dir, "ukb48298.tab.gz")
tg_snps.file <- paste0(data.dir, "tg_anthro_vignette_variants.txt")
pvar.file <- paste0(data.dir, "ewis_genotypes.pvar")
pgen.file <- paste0(data.dir, "ewis_genotypes.pgen")
psam.file <- paste0(data.dir, "ewis_genotypes.psam")
bmi_ss.file <- paste0(data.dir, "21001_EUR")

# output files
res.file <- paste0(res.dir, "ukbb_vqtl_tg_gxe_lmresults_20210908.csv")

# Load data
dd.df <- dd.file  %>%
  fread(data.table = F, stringsAsFactors = F)
dd <- dd.df %>%
  filter(FieldID %in% tg_fields)
non_qc_fields <- dd %>% 
  select(Field, FieldID) %>%
  filter(!grepl("QC", Field))
phe.df <- phe.file %>%
  fread(data.table = F, stringsAsFactors = F)#, nrows = 10000)
exp.df <- exp.file %>%
  fread(data.table = F, stringsAsFactors = F)#, nrows = 10000)
var_info.df <- var_info.file %>%
  fread(data.table = F, stringsAsFactors = F)
pc.df <- pc.file %>%
  fread(data.table = F, stringsAsFactors = F) %>%
  select(id, PC1, PC2) %>%
  rename(anthro_PC1 = PC1, anthro_PC2 = PC2)
whr.df <- whr.file %>%
  fread(data.table = F, stringsAsFactors = F)
nmr.df <- nmr.file %>%
  fread(data.table = F, stringsAsFactors = F)#, nrows = 10000)

# filter to anthro fields
anthro <- subset(var_info.df, var_info.df$Category_Level_3=="Anthropometry")
exp.df = exp.df[,(names(exp.df) %in% c("id",anthro$FieldID))]

# load genetics 
tg_snps.df <- tg_snps.file %>%
  fread(data.table = F, stringsAsFactors = F, header = F)
pvar.df <- pvar.file %>%
  fread(data.table = F, stringsAsFactors = F)
var_ids <- match(tg_snps.df$V1, pvar.df$ID)

psam.df <- psam.file %>%
  fread(data.table = F, stringsAsFactors = F)
pvar <- pgenlibr::NewPvar(pvar.file)
pgen <- pgenlibr::NewPgen(pgen.file, pvar=pvar) #,sample_subset = c(1,2,3,4) )
tg_genotypes.df <- ReadList(pgen, var_ids, meanimpute=F)

geno.df <- cbind(psam.df, data.frame(tg_genotypes.df)) %>%
  select(-`#FID`, -SEX)
colnames(geno.df)[2:10] <- tg_snps.df$V1

bmi_ss.df <- bmi_ss.file %>% 
  fread(data.table = F, stringsAsFactors = F) %>%
  filter(SNPID %in% tg_snps.df$V1) %>%
  select(SNPID, Non_Effect_Allele, Effect_Allele, `Beta_G-e`, `Var_Beta_G-e`) %>%
  group_by(SNPID) %>%
  mutate(pos_effect_allele = ifelse(`Beta_G-e` > 0, Effect_Allele, Non_Effect_Allele),
         pos_beta_ge = abs(`Beta_G-e`))

# subset nmr data
nmr_baseline.df <- nmr.df[, c(1, which(names(nmr.df) %in% c(tg_baseline_fields,process_fields)))]

# remove values where QC field is either 1 3 or 9
for (qi in 1:length(qc_map)){
  qc_field <- as.character(qc_map[qi])
  field <- names(qc_map[qi])
  if (field == 'f.23557.0.0') next
  to_remove <- which(nmr_baseline.df[[qc_field]] %in% c(1,3,9))
  nmr_baseline.df[to_remove, field] <- NA
}

# generate inv norm
for (f in names(nmr_baseline.df)){
  field_id <- unlist(strsplit(f, "\\."))[2]
  if (field_id %in% non_qc_fields$FieldID){
    nmr_baseline.df[!is.na(nmr_baseline.df[,f]), paste0(f, "_adj")] <- RNOmni::RankNorm(nmr_baseline.df[!is.na(nmr_baseline.df[,f]),f])  
  }
}

# merge it all together
data.df <- merge(phe.df, exp.df, by = "id") %>%
  merge(pc.df, by = 'id') %>%
  merge(whr.df, by.x = 'id', by.y = 'f.eid') %>%
  merge(nmr_baseline.df, by.x = 'id', by.y = 'f.eid') %>%
  merge(geno.df, by.x = 'id', by.y = 'IID')

nmr_baseline.df <- nmr_baseline.df %>%
  filter(f.eid %in% !!data.df$id)

fwrite(data.df, file = paste0(res.dir, "ukbb_vqtl_tg_gxe_inputdata_20210908.csv"), sep = ",")

```


```{r, include = F}

stats.df <- list()
for (f in names(nmr_baseline.df)){
  if ((f == 'f.eid') | (endsWith(f, "adj"))){
    next
  } else {
    field_id <- unlist(strsplit(f, "\\."))[2]
    field_name <- dd[dd$FieldID == field_id, "Field"]  
  }
  
  if (field_id %in% non_qc_fields$FieldID){
    stats.df[[f]] <- data.frame(
      field = field_name,
      n = sum(!is.na(nmr_baseline.df[[f]])),
      mean = mean(nmr_baseline.df[[f]], na.rm = T),
      min = min(nmr_baseline.df[[f]], na.rm = T),
      max = max(nmr_baseline.df[[f]], na.rm = T),
      std = sd(nmr_baseline.df[[f]], na.rm = T)
    )
  }
  
}

stats.df <- do.call(rbind, stats.df)


```

\newpage
# Raw field overview

```{r, echo = F}

knitr::kable(stats.df, digits=3, row.names = F)

```

\newpage
# Raw field distributions

```{r, fig.show="hold", out.width="50%", echo = F, message=FALSE}
cols <- rep(ggsci::pal_jco()(6),3)
ci <- 1
for (fi in 1:length(plot_fields)){
  f <- plot_fields[fi]
  field_id <- unlist(strsplit(f, "\\."))[2]
  field_name <- dd[dd$FieldID == field_id, "Field"]
  plt <- ggdensity(nmr_baseline.df, x = f, title = field_name, fill = cols[ci])
  ci <- ci +1
  suppressWarnings(print(plt))
}


```

\newpage
# Inv norm field distributions

```{r, fig.show="hold", out.width="50%", echo = F, message=FALSE}
cols <- rep(ggsci::pal_jco()(6),3)
ci <- 1
for (fi in 1:length(plot_adj_fields)){
  f <- plot_adj_fields[fi]
  field_id <- unlist(strsplit(f, "\\."))[2]
  field_name <- dd[dd$FieldID == field_id, "Field"]
  
  plt <- ggdensity(nmr_baseline.df, x = f, title = paste(field_name, "(norm)"), fill = cols[ci])
  ci <- ci +1
  suppressWarnings(print(plt))
}


```

\newpage
# Model

lm(metab ~ snp*exposure + sex + age + PC1-10)

```{r,  include=F}



# #interaction model I used in WGHS (note that it does not include sex in an all-women study, but should for UKB;
# exposures <- c("bmi","anthro_PC1", "anthro_PC2")
exposures <- c("anthro_PC1", "anthro_PC2", "WHRadjBMI_INT", names(exp.df)[-1])
# exposures <- names(exp.df)[-1]


final_df <- list()
for (snpi in 1:nrow(tg_snps.df)){
  snp <- tg_snps.df$V1[snpi]
  for (exposure in exposures){
    if (exposure %in% dd.df$FieldID){
      exposure_name <- dd.df[dd.df$FieldID == exposure, "Field"]
    } else {
      exposure_name <- exposure
    }
    for (metab in plot_adj_fields){
      metab_name <- dd.df[dd.df$FieldID == unlist(strsplit(metab, "\\."))[2], "Field"]
      fit <- lm(data.df[,metab] ~ data.df[,snp]*data.df[,exposure] + sex + age + PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10 + f.23649.0.0 + f.23650.0.0, data=data.df, na.action=na.exclude)
      res <- data.frame(summary(fit)$coefficients)
      final_df[[length(final_df)+1]] <- data.frame(
        outcome = metab_name,
        snp = snp,
        effect_allele = pvar.df[var_ids[snpi],"ALT"],
        exposure = exposure_name,
        snp_beta = res[2,1],
        snp_se = res[2,2],
        snp_p = res[2,4],
        exposure_beta = res[3,1],
        exposure_se = res[3,2],
        exposure_p = res[3,4],
        gxe_beta = res[nrow(res),1],
        gxe_se = res[nrow(res),2],
        gxe_p = res[nrow(res),4],
        stringsAsFactors = F
      )
      if (final_df[[length(final_df)]]$effect_allele != bmi_ss.df[bmi_ss.df$SNPID == snp, 'pos_effect_allele']){
        final_df[[length(final_df)]]$effect_allele <- bmi_ss.df[bmi_ss.df$SNPID == snp, 'pos_effect_allele']
        final_df[[length(final_df)]]$snp_beta <- -final_df[[length(final_df)]]$snp_beta
        final_df[[length(final_df)]]$gxe_beta <- -final_df[[length(final_df)]]$gxe_beta
      }

      # snprow=data.frame(as.list(summary(fit)$coefficients[c(2),c(1,2,4)]))
      # exprow=data.frame(as.list(summary(fit)$coefficients[c(3),c(1,2,4)]))
      # gxerow=data.frame(as.list(summary(fit)$coefficients[c(16),c(1,2,4)]))
      # cbind(metab, snp, exposure, snprow, exprow, gxerow)
    }
  }
}

final.df <- do.call(rbind, final_df)
fwrite(final.df, file = res.file, sep = ",")

```


```{r, echo = F}
print(paste("Model outputs written to:", res.file))

```