---
output: html_document
title: "Phenotype QC for UK Biobank vQTL analysis"
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F, message=F,warning=F, 
                      fig.path="../output/1b_vqtl_phenos_summary/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "cowplot"), 
  library, character.only=T))
```

```{r read-phenos}
raw_phenos <- read_csv("../data/processed/vqtl_phenos_raw.csv")
eur_phenos <- read_csv("../data/processed/vqtl_phenos_EUR.csv")
afr_phenos <- read_csv("../data/processed/vqtl_phenos_AFR.csv")
eas_phenos <- read_csv("../data/processed/vqtl_phenos_EAS.csv")
sas_phenos <- read_csv("../data/processed/vqtl_phenos_SAS.csv")
```

## Exclusion criteria

* No diabetes (self-report of diagnosed diabetes; field 2443)
* No coronary heart disease (self-report of diagnosed MI or angina; field 6150)
* No liver cirrhosis (ICD-10 codes K70.2-4, K71.7, K74.X; field 41202 (primary) or 41204 (secondary))
* No end-stage renal disease (algorithmically-defined; field 42026)

```{r exclusion-criteria}
raw_phenos_anc <- raw_phenos %>%
  mutate(ancestry = case_when(
    id %in% eur_phenos$id ~ "EUR",
    id %in% afr_phenos$id ~ "AFR",
    id %in% eas_phenos$id ~ "EAS",
    id %in% sas_phenos$id ~ "SAS",
    TRUE ~ as.character(NA)
  )) %>%
  filter(!is.na(ancestry))

raw_phenos_anc %>% 
  group_by(ancestry, diabetes) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = "diabetes", values_from = "n") %>%
  kable(caption = "Diabetes status") %>%
  kable_styling(full_width = F)

raw_phenos_anc %>% 
  group_by(ancestry, CHD) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = "CHD", values_from = "n") %>%
  kable(caption = "CHD status") %>%
  kable_styling(full_width = F)

raw_phenos_anc %>% 
  group_by(ancestry, cirrhosis) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = "cirrhosis", values_from = "n") %>%
  kable(caption = "Cirrhosis status") %>%
  kable_styling(full_width = F)

raw_phenos_anc %>% 
  group_by(ancestry, esrd) %>%
  summarise(n = n()) %>%
  pivot_wider(names_from = "esrd", values_from = "n") %>%
  kable(caption = "ESRD status") %>%
  kable_styling(full_width = F)

raw_phenos_anc %>%
  filter(!diabetes, !CHD, !cirrhosis, !esrd) %>%
  group_by(ancestry, is.na(alt_adj)) %>%
  summarise(n = n()) %>%
  kable(caption = "Sample sizes after exclusions") %>%
  kable_styling(full_width = F)
```

## Biomarker preprocessing for vQTL analysis

```{r biomarkers}
ukb_biomarker_fields <- c(
  alt = 30620, alb = 30600, alp = 30610, apoA = 30630, apoB = 30640, 
  ast = 30650, hscrp = 30710, Ca = 30680, chol = 30690, creatinine = 30700, 
  cysC = 30720, bilirubin_dir = 30660, ggt = 30730, glu = 30740, hba1c = 30750, 
  hdl = 30760, igf1 = 30770, ldl = 30780, lipA = 30790, oestradiol = 30800, 
  phos = 30810, rheum_factor = 30820, shbg = 30830, tes = 30850, 
  bilirubin_tot = 30840, protein_tot = 30860, tg = 30870, urate = 30880, 
  urea = 30670, vitD = 30890 
)

biomarkers <- names(ukb_biomarker_fields)
biomarkers_clean <- c(
  "Alanine aminotransferase", "Albumin", "Alkaline phosphatase", 
  "Apolipoprotein A", "Apolipoprotein B", "Aspartate aminotrasferase", 
  "High-sensitivity C-reactive protein", "Calcium", "Total cholesterol", 
  "Creatinine", "Cystatin C", "Direct bilirubin", "Gamma-glutamyl transferase", 
  "Random glucose", "Glycated hemoglobin", 
  "High-density lipoprotein cholesterol", "Insulin-like growth factor 1", 
  "Low-density lipoprotein cholesterol", "Lipoprotein A", "Estradiol", 
  "Phosphate", "Rheumatoid factor", "SHBG", "Testosterone", "Total bilirubin", 
  "Total protein", "Triglycerides", "Urate", "Urea", "Vitamin D"
)
biomarkers_short <- c(
  "ALT", "Alb", "ALP", "ApoA", "ApoB", "AST", "hsCRP", "Ca", "TC", "Creatinine",
  "CysC", "DBIL", "GGT", "RG", "HbA1c", "HDL-C", "IGF-1", "LDL-C", "LipA",
  "E2", "Phos", "RF", "SHBG", "Tes", "TBIL", "TPROT", "TG", "Urate", "Urea", 
  "VitD"
)
biomarker_units <- c(
  "U/L", "g/L", "U/L", "g/L", "g/L", "U/L", "mg/L", "mmol/L", "mmol/L", "umol/L",
  "mg/L", "umol/L", "U/L", "mmol/L", "mmol/mol", "mmol/L", "nmol/L", "mmol/L",
  "nmol/L", "pmol/L", "mmol/L", "IU/mL", "nmol/L", "nmol/L", "umol/L", "g/L",
  "mmol/L", "umol/L", "mmol/L", "nmol/L"
)
metabolic <- c(
  T, T, T, T, T, T, T, F, T, T, 
  F, T, T, T, T, T, F, T, T, F, 
  F, F, F, F, T, F, T, F, T, F
)

analytical_ranges <- list(
  alt = c(3, 500), alb = c(15, 60), alp = c(5, 1500), apoA = c(0.4, 2.5), apoB = c(0.4, 2), 
  ast = c(3, 1000), hscrp = c(0.08, 80), Ca = c(1, 5), chol = c(0.5, 18), creatinine = c(0, 4420), 
  cysC = c(0.1, 8), bilirubin_dir = c(0, 171), ggt = c(5, 1200), glu = c(0.6, 45), hba1c = c(0, 0), 
  hdl = c(0.05, 4.65), igf1 = c(1.3, 195), ldl = c(0.26, 10.3), lipA = c(5.76, 189), oestradiol = c(73, 17621), 
  phos = c(0.32, 6.4), rheum_factor = c(10, 120), shbg = c(0.33, 233), tes = c(0.35, 55.52), 
  bilirubin_tot = c(0, 513), protein_tot = c(30, 120), tg = c(0.1, 11.3), urate = c(89, 1785), 
  urea = c(0.8, 50), vitD = c(10, 375) 
)

tibble(
  Biomarker = biomarkers_clean,
  Abbreviation = biomarkers_short,
  Units = biomarker_units,
  `UKB Field` = ukb_biomarker_fields,
  Metabolic = ifelse(metabolic, "Yes", "No")
) %>%
  kable(caption = "Blood biomarkers measured in the UK Biobank") %>%
  kable_styling(full_width = F)
```

Following the approach of Wang et al. 2013 (modified to conduct the *entire* process separately in males and females):

1. Take residuals from linear model adjusting for age and 10 genetic PCs
2. Remove outliers (>5 SDs from the mean)
3. Standardize residuals (mean zero, variance one)

### Raw biomarker distributions

```{r raw-distributions}
for (bm in biomarkers) {
  raw_hist <- ggplot(raw_phenos, aes_string(x=bm)) +
    geom_histogram() +
    geom_vline(xintercept=analytical_ranges[[bm]][1], linetype="dashed") +
    geom_vline(xintercept=analytical_ranges[[bm]][2], linetype="dashed") +
    labs(x = biomarkers_short[match(bm, biomarkers)],
         y = "Sample size")
  print(raw_hist)
}
```

### Processed biomarker distributions

```{r distributions}
merged_phenos <- bind_rows(list(
  EUR = eur_phenos,
  AFR = afr_phenos,
  EAS = eas_phenos,
  SAS = sas_phenos
), .id = "ancestry")

for (bm in biomarkers) {
  # plt_raw <- ggplot(eur_phenos, aes_string(x=bm)) +
  #   geom_histogram()
  # plt_adj_eur <- ggplot(eur_phenos, aes_string(x=paste0(bm, "_adj"))) +
  #   geom_histogram()
  # plt_adj_afr <- ggplot(afr_phenos, aes_string(x=paste0(bm, "_adj"))) +
  #   geom_histogram()
  # plt_adj_eas <- ggplot(eas_phenos, aes_string(x=paste0(bm, "_adj"))) +
  #   geom_histogram()
  # plt_adj_sas <- ggplot(sas_phenos, aes_string(x=paste0(bm, "_adj"))) +
  #   geom_histogram()
  # print(plot_grid(plt_adj_eur, plt_adj_afr, plt_adj_eas, plt_adj_sas, 
  #                 nrow = 2, labels = c("EUR", "AFR", "EAS", "SAS"),
  #                 label_x = 0.75, label_y = 0.95, label_fontface = "plain"))
  anc_hist <- ggplot(merged_phenos, aes_string(x=paste0(bm, "_adj"))) +
    geom_histogram() +
    labs(x = biomarkers_short[match(bm, biomarkers)],
         y = "Sample size") + 
    facet_wrap(~ancestry, scales = "free_y") 
  print(anc_hist)
}
```

### Test the 5 SD outlier removal threshold

```{r outlier-threshold}
SD_vec <- 3:7

# Test ALT
alt_thresh_vec <- with(eur_phenos, mean(alt, na.rm=T) + SD_vec * sd(alt, na.rm=T))
alt_thresh_tbl <- tibble(`Threshold (# SDs)` = SD_vec, 
                     `# samples removed` = as.integer(NA))
alt_thresh_plt <- eur_phenos %>%
  ggplot(aes(x=alt)) +
  geom_histogram()
for (i in 1:5) {
  color <- if (i == 3) "darkred" else "black"
  alt_thresh_plt <- alt_thresh_plt + 
    geom_vline(xintercept=alt_thresh_vec[i], color=color)
  num_excl <- sum(
    with(eur_phenos, findInterval(
      alt, mean(alt, na.rm=T) + c(-1, 1) * SD_vec[i] * sd(alt, na.rm=T)) != 1),
    na.rm=T
  )
  alt_thresh_tbl[alt_thresh_tbl$`Threshold (# SDs)` == SD_vec[i], 2] <- num_excl
}
alt_thresh_plt
kable(alt_thresh_tbl) %>%
  kable_styling(full_width=F)

# Test TG
tg_thresh_vec <- with(eur_phenos, mean(tg, na.rm=T) + SD_vec * sd(tg, na.rm=T))
tg_thresh_tbl <- tibble(`Threshold (# SDs)` = SD_vec, 
                     `# samples removed` = as.integer(NA))
tg_thresh_plt <- eur_phenos %>%
  ggplot(aes(x=tg)) +
  geom_histogram()
for (i in 1:5) {
  color <- if (i == 3) "darkred" else "black"
  tg_thresh_plt <- tg_thresh_plt + 
    geom_vline(xintercept=tg_thresh_vec[i], color=color)
  tg_thresh_tbl[tg_thresh_tbl$`Threshold (# SDs)` == SD_vec[i], 2] <- sum(
    with(eur_phenos, findInterval(
      tg, mean(tg, na.rm=T) + c(-1, 1) * SD_vec[i] * sd(tg, na.rm=T)) != 1),
    na.rm=T
  )
}
tg_thresh_plt
kable(tg_thresh_tbl) %>%
  kable_styling(full_width=F)
```

